---
title: "Quantium_task1"
author: "Keith.Dang"
date: "2024-05-28"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---
```{r setup, include=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cran.r-project.org"))

# Load necessary libraries
if (!require("data.table")) {
  install.packages("data.table")
  library(data.table)
}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```

```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options)
{
# this hook is used only when the linewidth option is not NULL
if (!is.null(n <- options$linewidth))
{
x = knitr:::split_lines(x)
# any lines wider than n should be wrapped
if (any(nchar(x) > n))
x = strwrap(x, width = n)
x = paste(x, collapse = "\n")
}
hook_output(x, options)
})
```
## R Markdown

## Load required libraries and datasets
Note that you will need to install these libraries if you have never used these
before.
```{r 0 Load Libraries, results = 'hide'}
#### Install packages
install.packages("data.table")
install.packages("ggplot2")
install.packages("ggmosaic")
install.packages("readr")
install.packages("dplyr")
#### Load required libraries
library(data.table)
library(ggplot2)
library(ggmosaic)
library(readr)
library(dplyr)
```

#### Point the filePath to where the datasets are located
#### assign the data files to data.tables
```{r}
filePath <- "C:/Users/DELL/Desktop/Intern Materials/quantium/"
transactionData <- fread(paste0(filePath,"QVI_transaction_data.csv"))
customerData <- fread(paste0(filePath,"QVI_purchase_behaviour.csv"))
View(transactionData)
View(customerData)
```

## Exploratory data analysis
The first step in any analysis is to first understand the data. Let's take a look
at each of the datasets provided.

### Examining transaction data
We can use `str()` to look at the format of each column and see a sample of the
data. As we have read in the dataset as a `data.table` object, we can also run
`transactionData` in the console to see a sample of the data or use
`head(transactionData)` to look at the first 10 rows.

```{r Examining transaction data}
#### Examine transaction data
str(transactionData)
glimpse(transactionData)
```
We can see that the date column is in an integer format. Let's change this to a
date format.

```{r Convert DATE to date format}
#### Convert DATE column to a date format
#### A quick search online tells us that CSV and Excel integer dates begin on 30 Dec 1899
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```

Check that we are looking at the right products by examining PROD_NAME
```{r}
typeof(transactionData[,PROD_NAME])
summary(transactionData$PROD_NAME)
```

The product names in PROD_NAME column contain string text combined brand name and pack size

Looks like we are definitely looking at potato chips but how can we check that
these are all chips? We can do some basic text analysis by summarising the
individual words in the product name.

```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries
#### such as products that are not chips
productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), "
")))
setnames(productWords, 'words')
View(productWords)
```

As we are only interested in words that will tell us if the product is chips or
not, let's remove all words with digits and special characters such as '&' from our
set of product words. We can do this using `grepl()`.

```{r}
# Remove digits, and special characters, and then sort the distinct words by frequency of occurrence.
#### Removing digits and special characters using grepl()
productWords[,incorrect:=grepl("[[:digit:][:punct:]]",words)]
productWords <- productWords[incorrect==FALSE,][,incorrect:=NULL] 
productWords <- productWords[words !=""]
head(productWords)
#### Let's look at the most common words by counting the number of times a word appears and
#### sorting them by this frequency in order of highest to lowest frequency
```

There are salsa products in the dataset but we are only interested in the chips
category, so let's remove these.

```{r}
#### Remove salsa products
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```

Next, we can use `summary()` to check summary statistics such as mean, min and max
values for each feature to see if there are any obvious outliers in the data and if
there are any nulls in any of the columns (`NA's : number of nulls` will appear in
the output if there are any nulls).

```{r initial summary}
#### Summarise the data to check for nulls and possible outliers
summary(transactionData)
```

There are no nulls in the columns but product quantity appears to have an outlier
which we should investigate further(the max value is 200, but the mean still around 2). Let's investigate further the case where 200 packets of chips are bought in one transaction.
```{r}
#### Filter the dataset to find the outlier
filter(transactionData, PROD_QTY == 200)
```

There are two transactions where 200 packets of chips are bought in one transaction
and both of these transactions were by the same customer.

```{r}
#### Let's see if the customer has had other transactions
filter(transactionData, LYLTY_CARD_NBR == 226000)
```

It looks like this customer has only had the two transactions over the year and is
not an ordinary retail customer. The customer might be buying chips for commercial
purposes instead. We'll remove this loyalty card number from further analysis.
```{r}
#### Filter out the customer based on the loyalty card number
filter(transactionData, LYLTY_CARD_NBR == 226000)
transactionData <- transactionData[LYLTY_CARD_NBR!=226000]
#### Re-examine transaction data
transactionData[PROD_QTY == 200]
summary(transactionData)
```

That's better. Now, let's look at the number of transaction lines over time to see
if there are any obvious data issues such as missing data.

```{r}
#### Count the number of transactions by date
transactionData %>% group_by(DATE) %>% count()
```
```{R}
##### or:
transactionData[,.N,DATE][order(DATE)]
```

There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's
create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a
chart of number of transactions over time to find the missing date.

```{r}
####  create a column of dates that includes every day from 1 Jul 2018 to 30 Jun 2019, and join it onto the data to fill in the missing day.
dates <- seq(as.Date("2018-07-01"), as.Date("2019-06-30"), by = "day") 
dates_dt <- data.table(DATE = dates)
transactions_by_day <- merge(transactionData, dates_dt, by="DATE", all.y =TRUE )
transactions_by_day[, N:=.N , DATE]
View(transactions_by_day)
```


```{r fig.width = 10, fig.height = 8, fig.align = "center"}
#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))
#### Plot transactions over time
ggplot(transactions_by_day, aes(x = DATE, y = N)) +
geom_line() +
labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
scale_x_date(breaks = "1 month") +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that there is an increase in purchases in December and a break in late
December. Let's zoom in on this.

```{r fig.width = 10, fig.height = 8, fig.align = "center"}
#### Filter to December and look at individual days
december_trans <- transactions_by_day[month(DATE) == 12, .(n = .N), by = .(DATE)]

ggplot(december_trans)+
  geom_line(mapping = aes(x=DATE, y=n))+
  labs(x="Day", y="Number of transactions", title ="Transactions in December")+
  scale_x_date(breaks = "1 day")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that the increase in sales occurs in the lead-up to Christmas and that
there are zero sales on Christmas day itself. This is due to shops being closed on
Christmas day.
Now that we are satisfied that the data no longer has outliers, we can move on to
creating other features such as brand of chips or pack size from PROD_NAME. We will
start with pack size.

```{r Create pack size}
#### Pack size
#### We can work this out by taking the digits that are in PROD_NAME
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]
#### Let's check if the pack sizes look sensible
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]
```

The largest size is 380g and the smallest size is 70g - seems sensible!

```{r fig.width = 10, fig.height = 8, fig.align = "center"}
#### Let's plot a histogram of PACK_SIZE since we know that it is a categorical variable and not a continuous variable even though it is numeric.
ggplot(transactionData, aes(x = PACK_SIZE)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Pack Sizes", x = "Pack Size", y = "Frequency")
```

Pack sizes created look reasonable.
Now to create brands, we can use the first word in PROD_NAME to work out the brand
name...
```{r Create brand name}
#### Brands. Create a column which contains the brand of the product, by extracting it from the product name.
transactionData[,BRAND:= tstrsplit(PROD_NAME," ")[[1]]]
#### Checking brands
head(transactionData)
```

Some of the brand names look like they are of the same brands - such as RED and
RRD, which are both Red Rock Deli chips. Let's combine these together.

```{r Clean brand names}
#### Clean brand names
transactionData[BRAND == "RED", BRAND := "RRD"]
#### I see some brand names incorrect, such as 
transactionData[BRAND=="Dorito",BRAND:="Doritos"][BRAND=="Infzns",BRAND:="Infuzions"][BRAND=="Smith",BRAND:="Smiths"][BRAND=="Snbts",BRAND:="Sunbites"][BRAND=="Red", BRAND:="RRD"]
#### Check again
transactionData %>% count(BRAND)
```
Or you can use:
```{r another way, eval=FALSE, include=FALSE}
#### Using function fcase() to execute the same task
transactionData[, BRAND := fcase(
  BRAND == "Dorito", "Doritos",
  BRAND == "Infzns", "Infuzions",
  BRAND == "Smith", "Smiths",
  BRAND == "Snbts", "Sunbites",
  default = BRAND
)]
```
There are 23 unique BRANDS in BRAND column

### Examining customer data
Now that we are happy with the transaction dataset, let's have a look at the
customer dataset.

```{r 1 Exploratory data analysis}
#### Examining customer data.Do some basic summaries of the dataset, including distributions of any key columns.
summary(customerData)
unique(customerData, by ="LIFESTAGE")
unique(customerData, by ="PREMIUM_CUSTOMER")
```

```{r}
#### Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
```

As the number of rows in `data` is the same as that of `transactionData`, we can be
sure that no duplicates were created. This is because we created `data` by setting
`all.x = TRUE` (in other words, a left join) which means take all the rows in
`transactionData` and find rows with matching values in shared columns and then
joining the details in these rows to the `x` or the first mentioned table.

Let's also check if some customers were not matched on by checking for nulls.
```{r Check for missing customer details}
#### Check duplicateds. See if any transactions did not have a matched customer.
sum(duplicated(data))
data<-data[!duplicated(data),]
#### Check for nulls
null_count<- data[,lapply(.SD, function(x) sum(is.na(x)))]
null_count
```

Great, there are no nulls! So all our customers in the transaction data has been
accounted for in the customer dataset.

Note that if you are continuing with Task 2, you may want to retain this dataset
which you can write out as a csv
```{r Code to save dataset as a csv}
fwrite(data, paste0(filePath,"QVI_data.csv"))
```
Data exploration is now complete!

## Data analysis on customer segments
Now that the data is ready for analysis, we can define some metrics of interest to
the client:
- Who spends the most on chips (total sales), describing customers by lifestage and
how premium their general purchasing behaviour is
- How many customers are in each segment
- How many chips are bought per customer by segment
- What's the average chip price by customer segment

We could also ask our data team for more information. Examples are:
- The customer's total spend over the period and total spend for each transaction
to understand what proportion of their grocery spend is on chips
- Proportion of customers in each customer segment overall to compare against the
mix of customers who purchase chips

Let's start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and
plotting the split by these segments to describe which customer segment contribute
most to chip sales.
```{r calculate total sales}
new_data<- data %>% 
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>% 
  summarise(totalsale=sum(TOT_SALES)) %>% 
  arrange(totalsale) %>% 
  as.data.table()
```
Or using:
```{r}
new_data <-data[,.(totalsale=sum(TOT_SALES)),c("LIFESTAGE","PREMIUM_CUSTOMER")][order(totalsale)]
```

```{r fig.width = 10, fig.height = 8, fig.align = "center"}
#### Total sales by LIFESTAGE and PREMIUM_CUSTOMER. Calculate the summary of sales by those dimensions and create a plot.
ggplot(data)+
  geom_bar(mapping = aes(x=LIFESTAGE,fill=PREMIUM_CUSTOMER))+
  labs(title = "Total Sales of Life Stages", subtitle="replied on Premium Customer")+
  theme(axis.text.x = element_text(size=10, angle = 45))
```
This chart above illustrates the general of each segment of customer. we can use the next chart to compare between particular customer segments
```{r fig.width = 10, fig.height = 8, fig.align = "center"}
ggplot(data, aes(x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(position = position_dodge())+
  labs(x = "Lifestage", y = "Average Units per Customer")+
  ggtitle("Average Units per Customer by Lifestage")+
  theme(axis.text.x = element_text(size=10, angle = 90))
```

Sales are coming mainly from Budget - older families, Mainstream - young
singles/couples, and Mainstream - retirees

Let's see if the higher sales are due to there being more customers who buy chips.

```{r}
#### Number of customers by LIFESTAGE and PREMIUM_CUSTOMER. Calculate the summary of number of customers by those dimensions and create a plot.
data[, .(number_customer = uniqueN(LYLTY_CARD_NBR)), by = .(LIFESTAGE, PREMIUM_CUSTOMER)][order(desc(number_customer))]
```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy
chips. This contributes to there being more sales to these customer segments but
this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer.
Let's have a look at this next.

```{r}
#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER.
average_units_per_cus <- data %>% 
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>% 
  summarise(totalsale = sum(TOT_SALES), unique_cus = n_distinct(LYLTY_CARD_NBR)) %>% 
  mutate(average_units_per_customer = totalsale / unique_cus) %>% arrange(desc(average_units_per_customer))
```
Or using:
```{r}
average_units_per_cus <-data[,.(totalsale=sum(TOT_SALES), unique_cus=n_distinct(LYLTY_CARD_NBR)),c("LIFESTAGE","PREMIUM_CUSTOMER")][,average_units_per_customer:=totalsale / unique_cus][order(-average_units_per_customer)]
```

Plotting the chart
```{r fig.width = 10, fig.height = 8, fig.align = "center"}
#### Plot the average number of units per customer by those two dimensions.
ggplot(average_units_per_cus, aes(weight=average_units_per_customer, x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(position = position_dodge())+
  labs(x = "Lifestage", y = "Average Units per Customer")+
  ggtitle("Average Units per Customer by Lifestage")+
  theme(axis.text.x = element_text(size=10, angle = 35))
```
Older families and young families in general buy more chips per customer

Let's also investigate the average price per unit chips bought for each customer
segment as this is also a driver of total sales.

```{r}
average_price <- data %>% 
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>% 
  summarise(totalprice = sum(TOT_SALES), totalchips=sum(PROD_QTY)) %>% 
  mutate(average_price_per_unit = totalprice / totalchips) %>% 
  arrange(desc(average_price_per_unit))
```
Or using:
```{r}
average_price <- data[, .(totalprice = sum(TOT_SALES),totalchips=sum(PROD_QTY)), c("LIFESTAGE","PREMIUM_CUSTOMER")][,average_price_per_unit:= totalprice / totalchips][order(-average_price_per_unit)]
```

Plotting the chart
```{r fig.width = 10, fig.height = 8, fig.align = "center"}
#### Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER. Plot the average price per unit sold (average sale price) by those two customer dimensions.
ggplot(average_price, aes(weight=average_price_per_unit, x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(position = position_dodge())+
  labs(x = "Lifestage", y = "Average Price per units")+
  ggtitle("Average Price per units by Lifestage")+
  theme(axis.text.x = element_text(size=10, angle = 35))
```

Mainstream midage and young singles and couples are more willing to pay more per
packet of chips compared to their budget and premium counterparts. This may be due
to premium shoppers being more likely to buy healthy snacks and when they buy
chips, this is mainly for entertainment purposes rather than their own consumption.
This is also supported by there being fewer premium midage and young singles and
couples buying chips compared to their mainstream counterparts.

As the difference in average price per unit isn't large, we can check if this difference is statistically different.

```{r}
#### Perform an independent t-test between mainstream vs premium and budget midage and
data[,AVR_PRICE:= TOT_SALES/PROD_QTY]
t.test(data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") & PREMIUM_CUSTOMER == "Mainstream", AVR_PRICE], 
      data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") & PREMIUM_CUSTOMER != "Mainstream", AVR_PRICE], alternative = "greater")
#### young singles and couples. Perform a t-test to see if the difference is significant.
```
The t-test results in a p-value of 2.2*10^-16, i.e. the unit price for mainstream,
young and mid-age singles and couples ARE significantly higher than
that of budget or premium, young and midage singles and couples.

## Deep dive into specific customer segments for insights
We have found quite a few interesting insights that we can dive deeper into.

We might want to target customer segments that contribute the most to sales to
retain them or further increase sales. Let's look at Mainstream - young
singles/couples. For instance, let's find out if they tend to buy a particular
brand of chips.
```{r the most product this segment bought}
product_segments<- data %>% 
  filter(LIFESTAGE=="YOUNG SINGLES/COUPLES"& PREMIUM_CUSTOMER=="Mainstream") %>% 
  group_by(BRAND) %>% 
  summarise(mostBrand=sum(PROD_QTY)) %>% arrange(desc(mostBrand))
View(product_segments)
```
It seems like the most Brand that Mainstream - young singles/couples bought is Kettle. 

```{r fig.align = "center"}
#### Deep dive into Mainstream, young singles/couples. Work out of there are brands that these two customer segments prefer more than others. You could use a technique called affinity analysis or a-priori analysis (or any other method if you prefer)
main_segment<- data[LIFESTAGE=="YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER=="Mainstream"]
other_segments <-data[!(LIFESTAGE=="YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER=="Mainstream")]
#### total number of product in each segment
tot_quantity_seg <- main_segment[,sum(PROD_QTY)]
tot_quantity_other<- other_segments[,sum(PROD_QTY)]
#### the rate of individual BRAND per total in each segment
rateofBrand_mainseg <- main_segment[, .(seg = sum(PROD_QTY) / tot_quantity_seg), by = BRAND]
rateofBrand_otherseg <- other_segments[,.(other=sum(PROD_QTY)/tot_quantity_other), BRAND]
#### The Ratio of main segment to others segment
inspect_bybrand <-merge(rateofBrand_mainseg,rateofBrand_otherseg)[,result:=seg/other][order(-result)]
View(inspect_bybrand)
```
This table(inspect_bybrand) indicates that Mainstream young singles/couples purchased
Tyrrells chips more 23% than the rest. Mainstream young singles/couples are 56% less likely to purchase Burger Rings compared to the rest.

The similar method to packsize of each segment
```{r}
####Preferred pack size compared to the rest of the population
rateofPack_seg <- main_segment[,.(seg=sum(PROD_QTY)/tot_quantity_seg), by=PACK_SIZE]
rateofPack_other<- other_segments[,.(other=sum(PROD_QTY)/tot_quantity_other), by=PACK_SIZE]

inspect_bypack <- merge(rateofPack_seg, rateofPack_other)[,result:=seg/other][order(-result)]
View(inspect_bypack)
```
It looks like Mainstream young singles/couples are 27% more likely to purchase a 270g pack of chips compared to the rest.

## Conclusion:
we can dedate from all information we explored and analyzed.
First, the revenues have mostly been due to Budget - older families, Mainstream - young singles/couples, and Mainstream retirees shoppers, in which we found Mainstream - young singles/couples and Mainstream - retirees buyers are more than the rest of the population. Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterpart. We’ve also found that Mainstream young singles and couples are 23% more likely to purchase Tyrrells chips compared to the rest of the population.

